# Mixed Effects

## Exercises

I assume you are using the project from the previous chapters. If you don't have
this, start from the 'Exploratory Data Visualisation' chapter.

Start a new script or markdown file and load the following libraries. Install
them if they are not available to you using `install.packages()`.
```{r}
library(tidyverse)

library(here)

# To fit mixed effects models
library(lme4)
library(lattice) # required for some plots in lme4.

# This package contains functions to simulate Simpson's paradox data.
library(bayestestR)

# set the ggplot theme. You might try some alternatives here.
# The documentation which appears if you type ?theme_bw will also show the
# other possible prebuilt themes.
theme_set(theme_bw())
```


As before, load the data with:
```{r}
big_dia <- read_csv(here('data', 'big_dia.csv'))
```

### Simpson's Paradox

We looked at ['Simpson's paradox'](https://en.wikipedia.org/wiki/Simpson%27s_paradox) in the workshop presentation. This paradox can
be used to motivate both control variables and random effects. We're looking
from the perspective of random effects.

First simulate some data. You can change `n` or `groups` as you like.
```{r}
simpson_data <- simulate_simpson(n = 100, groups = 8)
```

Here is a simple plot of the data with a simple linear regression using
`geom_smooth()`.
```{r}
simpson_data |> 
  ggplot(
    aes(
      x = V1,
      y = V2
    )
  ) +
  geom_point() + 
  labs(
    V1 = "explanatory variable",
    V2 = "response"
  ) +
  geom_smooth(method="lm", se=FALSE, linewidth=3)
```

::: webex-box

Modify the plot so each group is coloured differently and has its own linear
trend line.

:::

The plot you have made, if you succeeded in the exercise, shows multiple
independent regressions (one for each group). This is not the same as mixed
effects, which can pool information together across the different groups.


::: {.webex-box}

1. Fit a simple linear regression model predicting V2 by V1 using `lm()`. Plot
predictions for this model using the techniques introduced in the previous 
session. What is wrong with the simple linear regression?

2. Complete the following code block to fit a model of the Simpson's paradox data
with random intercepts for each group.
```{{r}}
simpson_fit <- lmer(
  formula = V2 ~ V1 + (1|???),
  data = ???
)
```

Run the code.

:::

```{r}
#| code-fold: true
#| code-summary: "Click here for the answer to question 2."
simpson_fit <- lmer(
  formula = V2 ~ V1 + (1|Group),
  data = simpson_data
)
```

We can generate predictions from this model for each line in the data.

::: {.webex-box}

Complete the following code block to generate predictions for each group using
the `predict()` function
```{{r}}
simpson_data <- simpson_data |> 
  mutate(
    re_predictions = ???
  )
```

Run the code.

:::

```{r}
#| code-fold: true
#| code-summary: "Click here for the answer"
simpson_data <- simpson_data |> 
  mutate(
    re_predictions = predict(simpson_fit)
  )
```

We now plot these predictions:
```{r}
simpson_data |> 
  ggplot(
    aes(
      x = V1, 
      y = V2,
      colour = Group
    )
  ) +
  geom_point() +
  # override part of the aesthetic mapping to get predictions into `geom_path`
  geom_path(
    aes(y = re_predictions)
  )
```

Letting the model know about the grouping structure, via random intercepts,
is enough to fix the problem with the simple linear regression. If you are 
interested, compare this to a multiple linear regression with `Group` as fixed
effect. 

### Word duration and frequency

Complex mixed effects models benefit from scaling and/or centring variables.
This is true both for humans and for the computer. For us, it can help with
interpretation of models. For the machine, it can help with the mathematics of
fitting the model.

Both use the `scale()` function. 

::: webex-box

Look at the documentation for `scale()`. 

1. Work out how to scale only, centre only, and do both using the function. 
What is the default behaviour?

2. Create two new variables in `big_dia`:
    1. `centred_frequency`, which centres `unigram.google.gb`.
    2. `scaled_duration`, which scales and centres `WordDuration`
    
3. Fit a model with `scaled_duration` as the response, `centred_frequency` and
`final` as fixed effects, and random intercepts for `Speaker` and 
`TargetOrthography` (i.e. the word).

4. Look at the summary and plot the overall model prediction. Note that you can
generate predictions for each `Speaker` and ignoring `Speaker` by using the
`re.form` argument to `predict()`. Say `re.form = NA` to ignore groups.

:::

```{r}
#| code-fold: true
#| code-summary: "Click here for the answer to (2)"
big_dia <- big_dia |> 
  mutate(
    centred_frequency = scale(
      unigram.google.gb, 
      scale=FALSE
    ),
  scaled_duration = scale(WordDuration)
  )
```


```{r}
#| code-fold: true
#| cache: true
#| code-summary: "Click here for the answer to (3)"
intercept_fit <- lmer(
  scaled_duration ~ centred_frequency + final +
    (1|Speaker) + (1|TargetOrthography),
  data = big_dia
)
```

::: webex-box

We can also add random slopes. Recall that `(1 + variable|group)` fits a 
random intercept _and slope_ for each group and estimates the
correlation of the intercepts and slopes.

5. Extend the model you fit in (3) to include a random slope on 
`centred_frequency`. Explore this model using the tools we have considered
already.
:::

```{r}
#| code-fold: true
#| code-summary: "Click here for the answer to (5)"
slope_fit <- lmer(
  scaled_duration ~ centred_frequency + final +
    (1 + centred_frequency|Speaker) + (1|TargetOrthography),
  data = big_dia
)
```

::: webex-box

How do we test the significance of `centred frequency`? We discussed using
the `anova()` function for model comparison. 

6. Fit a model which matched the model in (5), but doesn't include 
`centred_frequency` anywhere. Use the argument `REML = FALSE` in `lmer()`. 
Compare this model with the model in (5) using the `anova()` function, with argument `test = 'Chisq'`. Does this method indicate a statistically 
significant effect of `centred_frequency`? What is the $p$-value?

We can extracting random effects
terms from our model using the `ranef()` function.

```{r}
re_slope <- ranef(slope_fit)
```

7. Look at the structure of this object in the RStudio Environment pane (top
right)

8. Make a histogram of the random intercepts of `TargetOrthography`. Does this look normally
distributed to you?

9. Create a scatter plot of speaker intercepts and slopes for `Speaker`. Are
these values correlated? What might this suggest about the effect of 
frequency on duration?

:::


